To be FAIR: Theory Specification Needs an Update

### 2025-06-07 action_letter.Rmd
update 2025-06-07 15:14:35.120732

```{asis}
The second kind of theorizing I have done is more philosophical, along the lines of my "Three Laws" paper and related work. Here, I think applying the FAIR framework would be a little more problematic. At times, the current manuscript sounds as though it is trying to make scientific theory machine readable, an idea that has some good points that the authors make clear, but would also eliminate a lot of interesting papers that have been written by human beings from a subjective point of view. What would happen to Meehl's great Sir Karl Sir Ronald paper within FAIR? How much of the importance of that paper would remain if you took Meehl's voice out of it? Is all narrative philosophy "theorizing" that could be augmented or replaced by FAIR? See also: Meehl on "Cliometric metatheory".
```

Discuss relationship between FAIR and prose/author's voice.

@@

```{asis}
These considerations lead me to some recommendations for strengthening the manuscript. The first is to clarify the relationship between FAIR theory and traditional paper-writing. Is the former meant to replace or enhance the latter? It seems to me that using FAIR to enhance the environment created by traditional theoretical papers is much the better option.  That is, everything that currently exists about the S-R hypothesis could continue to exist, and be supplemented by a FAIR framework outlining the theory and its implications. Development of the FAIR model could exist almost independently of the individual papers, analyses, reviews and meta-analyses (see below). Someone could contribute to the literature on a subject by developing a FAIR framework, much as a good review paper does.
```

Yes, this is absolutely how we intended it. Clarify this.

@@

This is not exactly true; while that's definitely in line with *my* usage of FAIR theory, other uses are possible.

>>>>

```{asis}
The second kind of theorizing I have done is more philosophical, along the lines of my "Three Laws" paper and related work. Here, I think applying the FAIR framework would be a little more problematic. At times, the current manuscript sounds as though it is trying to make scientific theory machine readable, an idea that has some good points that the authors make clear, but would also eliminate a lot of interesting papers that have been written by human beings from a subjective point of view. What would happen to Meehl's great Sir Karl Sir Ronald paper within FAIR? How much of the importance of that paper would remain if you took Meehl's voice out of it? Is all narrative philosophy "theorizing" that could be augmented or replaced by FAIR? See also: Meehl on "Cliometric metatheory".

These considerations lead me to some recommendations for strengthening the manuscript. The first is to clarify the relationship between FAIR theory and traditional paper-writing. Is the former meant to replace or enhance the latter? It seems to me that using FAIR to enhance the environment created by traditional theoretical papers is much the better option.  That is, everything that currently exists about the S-R hypothesis could continue to exist, and be supplemented by a FAIR framework outlining the theory and its implications. Development of the FAIR model could exist almost independently of the individual papers, analyses, reviews and meta-analyses (see below). Someone could contribute to the literature on a subject by developing a FAIR framework, much as a good review paper does.
```

We agree, this is exactly how we envision the role of FAIR theory. We think this comment can be concisely addressed by emphasizing that modular publishing does not replace the traditional paper, but adds value to resources created in the process of its writing:

`r get_revision("modularvalue")`

And:

@@

@@

We partly agree; we recognize two distinct lines of thought in this comment.

The first is that FAIR theory relates to construct validity. To address this point, we have rewritten the section on FAIR theory and the "measurement crisis":

`r get_revision("measurementcrisis")`

The second is that much of what we describe looks like Meehl's nomological nets. This is definitely true for the way that *we* are using FAIR theory (e.g., in the online examples), as we are inspired by Meehl,
but it's not necessarily true of FAIR theory in general.
We're not sure if further changes need to be made to address this.

### 2025-06-07 manuscript.Rmd
update 2025-06-07 15:14:35.120732

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments.
  This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta science, theory formation, open science"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "fair_theory.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, noextraspace"
urlcolor          : blue
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            :
  papaja::apa6_pdf: default
  md_document: default
header-includes:
  - \usepackage{longtable, setspace, subfig}
  - \raggedbottom
---

@@

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts.
<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
<!-- which is used by many academic publishers. -->
The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.

<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication.
</span>

@@

Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
Since FAIR theories can reference other theories and resources,
it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.
<span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,
which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct [@songLiteratureReviewingAddressing2021].
By explicitly referencing operational definitions in FAIR theories,
such jingle-jangle fallacies would come to light and could ultimately be resolved.</span>

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Conceptualization
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University department of Methodology & Statistics, Tilburg, The Netherlands"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, Amsterdam, The Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, München, Germany"
  - id            : "7"
    institution   : "MSB Medical School Berlin Department of Psychology, Berlin, Germany"
authornote: |
  We are particularly endebted to our Reviewer, Dr. Eric Turkheimer, whose thoughtful reflections on FAIR theory were incorporated into the Discussion. We also thank the Editor, Dr. Scott Vrieze, and two anonymous Reviewers for their detailed and invaluable comments. This version of the paper was generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.
abstract: "`r tryCatch(trimws(readr::read_file(here::here('abstract.md'))))`"
keywords          : "fairtheory, meta science, theory formation, open science"
wordcount         : "`r tryCatch(wordcountaddin::word_count(here::here('manuscript.Rmd')))`"
bibliography      : "fair_theory.bib"
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man, noextraspace"
urlcolor          : blue
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
documentclass     : "apa7"
output            :
  papaja::apa6_pdf: default
  md_document: default
header-includes:
  - \usepackage{longtable, setspace, subfig}
  - \raggedbottom
---

@@

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015].
These various practices ensure transparent and repeated testing of hypotheses by committing to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeachersCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

@@

We envision FAIR theory as an instance of modular publishing [@kirczModularityNextForm1998].
<span id = "modularvalue">The primary unit of scientific communication has long been the academic paper.
In the process of writing papers, scholars often produce many valuable resources, including instruments, materials, data, code, and theory.
These resources are often merely described in papers (if at all), and not made available for reuse.
Modular publishing is the practice of making each of these resources available as independent *information artifacts*,
facilitating their reuse and making them citable [@vandesompelRethinkingScholarlyCommunication2004].</span>
<span id="datasharing">Data sharing is an example of modular publishing that is already widely adopted, becoming the de-facto standard in psychology [@tedersooDataSharingPractices2021].</span>

Modular publishing can be achieved by archiving specific resources (including theory) in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts.
<!-- - as does, for example, the [Crossref](https://www.crossref.org/) association, -->
<!-- which is used by many academic publishers. -->
The DataCite Metadata Schema offers a standard way to document information artifacts with relevant metadata which increases their findability and documents their relationships to other resources [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, data can be archived in Zenodo with the metadata property `resourceType: dataset`.
If the data were collected for a specific paper, that relationship can be cross-referenced with `relationType: IsSupplementTo`.
Similarly, a FAIR theory can be connected to a specific paper by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`.
Other cross-references are useful for documenting relationships between multiple theory objects: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<span id="modularpublisingdetract">Modular publishing of resources, including theories, increases their reuse potential and makes them citable without detracting from the conventional academic paper as a unit of academic communication which allows for argumentation,
nuance,
and the author's voice.
</span>

@@

<span id="measurementcrisis">Another future direction is the intersection between the aforementioned "theory crisis" and the related "measurement crisis" pertaining to the lack of clarity, consistency, and validity in the operationalization of theoretical constructs [@bringmannBackBasicsImportance2022].
Since FAIR theories can reference other theories and resources,
it is possible to attach references to specific measurement instruments (or even theories of measurement) to constructs named in a theory.</span>
<span id="jinglejangle">FAIR theories can also help address "jingle- and jangle fallacies" in psychology,
which are ambiguities that arise from using the same term for different constructs, or conversely, using different terms for the same construct [@songLiteratureReviewingAddressing2021].
By explicitly referencing operational definitions in FAIR theories,
such jingle-jangle fallacies would come to light and could ultimately be resolved.</span>
